\documentclass[conference]{IEEEtran}

\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{url}
\usepackage{tikz}
\usepackage{listings}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,fit,calc,backgrounds,decorations.pathreplacing}

\definecolor{mralight}{rgb}{0.95,0.96,0.98}
\definecolor{mradark}{rgb}{0.10,0.18,0.28}
\lstset{
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  columns=fullflexible,
  frame=single,
  framesep=4pt,
  backgroundcolor=\color{mralight},
  xleftmargin=0.4em,
  xrightmargin=0.4em,
  framextopmargin=4pt,
  framexbottommargin=4pt,
  aboveskip=1.2em,
  belowskip=1.2em
}
% Add vertical space between a top listing caption and its colored box.
\makeatletter
\let\MRA@lstMakeCaption\lst@MakeCaption
\def\lst@MakeCaption#1{\MRA@lstMakeCaption{#1}\if t#1\vskip5\p@\relax\fi}
\makeatother

\newcommand{\mra}{Marine Race Arena}
\newcommand{\cmd}[1]{\texttt{#1}}

\begin{document}

\title{\mra: A Configurable HoloOcean Benchmark for Underwater Gate Racing and Team-Level Fleet Evaluation}

\author{\IEEEauthorblockN{Lorenzo Bacchiani, Andrea Bedei, Roberto Girau and Giovanni Pau}
\IEEEauthorblockA{Department of Computer Science and Engineering, University of Bologna, Italy}
}

\maketitle

\begin{abstract}
Autonomous racing provides a compact and reproducible setting for evaluating perception, planning and control under time constraints. Ground and aerial robotics have established racing platforms and gate-based competitions, but underwater robotics lacks a comparable benchmark for speed-oriented racing along predefined gate courses. This paper presents \mra{}, a configurable benchmark layer built on HoloOcean with a BlueROV2 vehicle. A single JSON configuration specifies the track, gates, sensors, currents, obstacles and scoring rules. The benchmark strictly separates control from evaluation. Controllers receive only an official observation composed of onboard sensing, acoustic guidance and race progress. An independent referee uses privileged simulator state only for gate validation, penalties, timeouts and scoring. We formalize the benchmark model, controller interface, gate-crossing rules and scoring procedure. We also extend the benchmark to a staggered multi-rover team with per-vehicle referee state and team-level scoring. In HoloOcean, a deterministic rule-based baseline completes the three official clean tracks and is also evaluated under optional disturbance currents. On a deterministic kinematic substrate, a second conservative controller differs measurably from the first in completion time and motion smoothness. A leader-follower controller uses only an optional realistic acoustic channel and reduces a staggered team's inter-vehicle proximity events to the single-rover level while every rover finishes. HoloOcean diagnostic validation on clean Horseshoe Bay tests this coordination with three rovers, no currents or obstacles, two start gaps and three seeds. Every coordinated run finishes with no inter-vehicle proximity events or gate/world collisions, whereas uncoordinated teams incur many gate/world collisions and sometimes fail to finish. The arena also supports optional obstacles. \mra{} provides a reproducible evaluation contract for developing and comparing autonomous underwater racing controllers, including future learning-based approaches.
\end{abstract}

\begin{IEEEkeywords}
underwater robotics, autonomous racing, HoloOcean, BlueROV2, simulation benchmark, referee, multi-robot systems, team-level evaluation
\end{IEEEkeywords}

\input{sections/01_introduction}
\input{sections/02_related_work}
\input{sections/03_system_overview}
\input{sections/04_environment_design}
\input{sections/05_controller}
\input{sections/06_referee}
\input{sections/07_fleet}
\input{sections/08_evaluation}
\input{sections/09_discussion}
\input{sections/10_conclusion}

\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}
