\section{Introduction}
\label{sec:introduction}

Autonomous racing has emerged as an effective research format for evaluating perception, planning, control and real-time decision-making within a single quantitative task under repeatable conditions~\cite{betz2022autonomous}. In ground robotics, the F1TENTH platform provides standardized vehicles, circuits, metrics and baselines for continuous control and reinforcement learning~\cite{okelly2020f1tenth}. At full scale, competitions such as the Abu Dhabi Autonomous Racing League (A2RL) push planning and control toward the limits of vehicle handling in head-to-head racing scenarios~\cite{hoffmann2026a2rl}. In aerial robotics, gate-based benchmarks and competitions have driven progress from modular perception and control pipelines~\cite{foehn2020alphapilot} to learning-based policies capable of champion-level drone racing~\cite{kaufmann2023swift}. Across these domains, the availability of a well-defined racing task, fixed course geometry, explicit scoring rules and machine-readable outcomes has made autonomy methods easier to compare, reproduce and improve.

\input{figures/task_teaser}

Underwater robotics has not yet benefited from an equivalent benchmark for speed-oriented autonomous racing along predefined gate courses. Existing marine simulators provide increasingly mature support for underwater vehicles, sensors and environments, but they mainly target intervention, inspection, navigation or sensor realism rather than racing as an evaluation protocol. In particular, they do not provide a compact underwater counterpart to ground and aerial racing benchmarks: a task defined by ordered gates, timing, penalties, currents, obstacles, independent scoring and reproducible outcomes. This gap matters because underwater racing is not simply aerial or ground racing transferred to a different simulator. Underwater vehicles are slow, strongly damped and affected by persistent currents; global localization is typically unavailable; inertial and Doppler sensing may drift; optical perception is limited in range and degraded by turbidity; acoustic cues are noisy and low bandwidth; and physical contact with gates, obstacles or other vehicles can be difficult to interpret. These properties change both the information available to a controller and the way in which performance should be evaluated.

A central requirement for such a benchmark is therefore a strict separation between autonomy and privileged simulation services. A controller should be evaluated using only information that would plausibly be available onboard, such as local sensing, physically received acoustic packets and controller-local state. Privileged simulator state, including global pose, exact gate geometry, official course progress and the true current field, must remain outside the controller boundary. The benchmark may use it internally to construct the physical environment and referee inputs, but official progress is evaluator-only and none of this state flows back to autonomy. Without this boundary, performance can depend on information that would not be available to a real vehicle, making results difficult to interpret and unfair to compare.

This paper presents \mra{}, a configurable benchmark layer for autonomous underwater gate racing (Fig.~\ref{fig:teaser}). Its benchmark model, controller contract and referee are vehicle-independent; replaceable adapters isolate simulator-specific sensing and actuation. The reference implementation and all reported experiments use HoloOcean~\cite{potokar2022holoocean} with a BlueROV2-class vehicle~\cite{vonbenzon2022bluerov2}. Rather than replacing a simulator or proposing an autonomy algorithm, \mra{} defines an evaluation contract on top of a simulation backend. A JSON configuration specifies the world, ordered gates, sensors, currents, obstacles and scoring rules. Controllers receive only an official observation, while an independent referee uses privileged state to validate crossings and violations, assign penalties, manage timeouts and compute official scores.

The core task first evaluates one autonomous vehicle whose controller is integrated through a small interface (Section~\ref{sec:controller}). Fleet mode then extends the same mechanism to a cooperative team in which each vehicle runs its own controller and retains independent progress. The separation from evaluation lets the benchmark compare arbitrary controllers, including future learning-based ones, under the same official observation.

The contributions of this paper are as follows.

\begin{enumerate}

\item We present Marine Race Arena, an end-to-end configurable benchmark for autonomous underwater gate racing. Users can define complete race scenarios, including the environment, vehicle configuration, sensor profile, ordered gates, currents, obstacles, timing and scoring through a single configuration file. The release includes three ready-to-use official tracks and supports the creation of custom courses and experimental scenarios.

\item We provide an extensible autonomy interface for controller development and comparison. The framework includes two camera-assisted rule-based reference controllers and allows users to integrate their own controller through a minimal three-method interface without modifying the race runner, simulator adapter or evaluation logic. A documented official observation contract ensures that every controller is evaluated using the same non-privileged information.

\item We introduce an independent automated referee that uses privileged simulator state only for evaluation. It validates ordered gate crossings, detects rule violations, applies penalties and timeouts, ranks participants and produces structured event logs and machine-readable race summaries.

\item We support both single-vehicle racing and cooperative multi-vehicle evaluation within the same framework. Each vehicle maintains independent control and a locally estimated course state, while the referee maintains separate official scoring state for the team result. The fleet layer provides staggered or simultaneous starts, inter-vehicle proximity diagnostics, optional acoustic communication and distributed leader-follower coordination.

\item We provide a reproducible experimental validation in HoloOcean covering the three official tracks, comparison between the two reference controllers, disturbance-current scenarios and multi-vehicle coordination. The accompanying configurations, scripts and structured outputs allow the reported experiments to be reproduced and extended with new tracks, vehicles and control strategies.

\end{enumerate}

The remainder of the paper is organized as follows. Section~\ref{sec:related_work} reviews underwater simulators, autonomous-racing benchmarks and multi-AUV underwater cooperation. Section~\ref{sec:system_overview} formalizes the benchmark model and runtime architecture. Section~\ref{sec:configuration} describes the vehicle, the HoloOcean adapter, the sensor suite, the current model and the official observation contract. Section~\ref{sec:controller} specifies the controller interface and the rule-based baseline. Section~\ref{sec:referee} formalizes gate validation, the referee state machine and single-vehicle scoring. Section~\ref{sec:fleet} describes staggered fleet execution and team scoring. Section~\ref{sec:evaluation} reports the experimental evaluation. Section~\ref{sec:discussion} discusses limitations and future work. Section~\ref{sec:conclusion} concludes the paper.
