\section{Benchmark Model and System Architecture}
\label{sec:system_overview}

We specify \mra{} as an \emph{evaluation problem} rather than a controller-design problem: the benchmark fixes the world, task and scoring, while a participant supplies only a policy. This section defines the benchmark instance, participant interface and runtime architecture that enforce the separation between the two.

\subsection{Benchmark Instance}

\noindent\textbf{Definition~1 (Benchmark instance).} A \mra{} instance is the tuple
\begin{equation}
\label{eq:instance}
\mathcal{E} = (\mathcal{W}, \mathcal{G}, \mathcal{S}, \mathcal{F}, \mathcal{C}, \mathcal{O}, \mathcal{P}, \mathcal{R}),
\end{equation}
where:
\begin{itemize}
 \item $\mathcal{W}$ is the \emph{world}: an axis-aligned bounds box $\mathcal{B}=[x_{\min},x_{\max}]\times[y_{\min},y_{\max}]\times[z_{\min},z_{\max}]\subset\mathbb{R}^3$ (with $z$ positive-up, so the navigable volume satisfies $z<0$) and the simulation environment selected through the adapter;
 \item $\mathcal{G}=(g_1,\dots,g_M)$ is the \emph{ordered} gate sequence; each gate is the tuple $g_k=(c_k,\,n_k,\,w_k,\,h_k)$ of center $c_k\in\mathbb{R}^3$, passage-direction vector $n_k\in\mathbb{R}^3$ (nonzero, not necessarily unit length) and inner aperture width $w_k$ and height $h_k$;
 \item $\mathcal{S}$ is the shared \emph{start default}: a base spawn pose and an optional release policy that a participant inherits only when it does not specify its own;
 \item $\mathcal{F}$ is the finish condition: completion of the last gate of the last lap (Section~\ref{sec:referee});
 \item $\mathcal{C}$ is the current field $\mathbf{v}_c:\mathbb{R}^3\times\mathbb{R}_{\ge0}\to\mathbb{R}^3$ (Section~\ref{sec:configuration});
 \item $\mathcal{O}$ is the obstacle set;
 \item $\mathcal{P}=\{1,\dots,N\}$ is the participant set; participant $i$ carries a vehicle, sensor profile and controller, together with an optional spawn pose and release delay $\tau_i\!\ge\!0$ that override the start default of $\mathcal{S}$;
 \item $\mathcal{R}$ is the referee configuration: gate-validation parameters, penalty values and scoring rules.
\end{itemize}
Each gate's orientation derives solely from its passage direction $n_g$, giving a right-handed frame
\begin{equation}
\label{eq:gate_frame}
\hat{n}=\frac{n_g}{\lVert n_g\rVert},\quad \hat{r}_g=\frac{\hat{z}\times\hat{n}}{\lVert\hat{z}\times\hat{n}\rVert},\quad \hat{s}_g=\hat{n}\times\hat{r}_g,
\end{equation}
with world-up $\hat{z}=(0,0,1)$; if the passage direction is vertical ($\lVert\hat{z}\times\hat{n}\rVert\!\approx\!0$), $\hat{r}_g$ falls back to the world $\hat{x}$ axis so the frame stays well defined. The gate basis is therefore unambiguous and independent of any visual rotation and is reused by the beacon placement (Section~\ref{sec:configuration}) and the crossing test (Section~\ref{sec:referee}).

\subsection{Participant Interface}

\noindent\textbf{Definition~2 (Policy).} A participant $i$ implements a policy
\begin{equation}
\label{eq:policy}
\pi_i: o_{i,t}\mapsto u_{i,t},
\end{equation}
mapping the official observation $o_{i,t}$ (Section~\ref{sec:configuration}) to a normalized, high-level, body-frame command
\begin{equation}
\label{eq:command}
u_{i,t}=\big[u^{\mathrm{surge}},\,u^{\mathrm{sway}},\,u^{\mathrm{heave}},\,u^{\mathrm{yaw}}\big]^{\!\top}\in[-1,1]^4 .
\end{equation}
The action of \eqref{eq:command} is the portable, constrained interface that an adapter maps to the selected vehicle. In fleet mode the policy may return $(u_{i,t},m_{i,t})$, where $m_{i,t}$ is an optional controller-authored communication payload and only $u_{i,t}$ is a physical action. The reference HoloOcean adapter also accepts a direct eight-thruster \cmd{thrusters} action for its BlueROV2-class agent. Time is discretized at a fixed control timestep $\Delta t$, with $t_k=k\,\Delta t$. The adapter clamps $u_{i,t}$ to actuator limits and applies depth-bound safety before mapping it to the vehicle; in official mode the policy reads only the official observation $o_{i,t}$, whose contract excludes privileged state (Section~\ref{sec:configuration}).

\subsection{Design Requirements}

The design follows the seven requirements in Table~\ref{tab:requirements}. They encode two commitments: scoring logic is independent of the participant and cannot be manipulated through privileged feedback; enabling fleet mode must not silently change single-vehicle behavior.

\begin{table}[t]
\centering
\caption{Design requirements for the benchmark layer.}
\label{tab:requirements}
\mratablestyle
\begin{tabular}{p{0.10\columnwidth}p{0.80\columnwidth}}
\toprule
ID & Requirement \\
\midrule
R1 & Race configuration is declarative and reproducible from a single set of track JSON files. \\
R2 & Official controller observations exclude ground-truth pose, simulator-only state and referee-derived course progress. \\
R3 & Referee logic is independent of the participant controller and cannot be influenced by it. \\
R4 & Gate dimensions, referee margins and track geometry are not changed to hide failures. \\
R5 & Every run emits structured event logs and machine-readable summaries. \\
R6 & Single-vehicle behavior is unchanged when fleet mode is disabled. \\
R7 & Fleet mode aggregates a team score while preserving per-vehicle diagnostics. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Runtime Architecture}

\input{figures/architecture_diagram}

The runtime separates scenario construction, vehicle autonomy and race evaluation. A declarative track configuration defines the environment, ordered gates, beacon network, disturbances, participants and scoring rules; after validation, it becomes an arena instance connected to a simulator adapter. The adapter isolates sensing, actuation and time advancement, allowing the controller and referee interfaces to be reused across simulation backends, while the engine-free adapter supports deterministic tests without contributing to the reported results. During execution, the race runner advances the simulator and mediates vehicle-environment interaction. The controller receives an observation containing only participant-local time, allow-listed onboard measurements, physically received beacon packets and optional inter-vehicle messages. It returns a bounded motion command and cannot access simulator state, gate geometry, race progress or referee decisions, so navigation and course progression must be inferred locally. In parallel, the referee uses privileged simulator state to validate crossings and rule violations and to produce event logs, participant results and team-level scores. Referee-derived targets, progress and status are never returned to the controller; disagreements with its local estimate are recorded for offline analysis rather than corrected. Figure~\ref{fig:runtime_architecture} summarizes this separation and the controller-visible observation boundary.
