\section{Experimental Evaluation}
\label{sec:evaluation}

This evaluation does not seek an optimal underwater racing controller. It tests whether the benchmark, official observation contract, referee and team scoring behave as specified, compares the two reference controllers across every reported condition, and reports where the reference baseline fails.

\subsection{Protocol and Metrics}

The final matrix contains 78 real-HoloOcean runs generated under one frozen source fingerprint recorded in the artifact manifest (Section~\ref{sec:reproducibility}). Clean single-vehicle validation crosses three tracks, two controllers and seeds 0--4 (30 runs). Current validation crosses the \texttt{medium} and \texttt{strong} Horseshoe Bay profiles, both controllers and the same five seeds (20 runs). Homogeneous two-vehicle fleet validation uses Horseshoe Bay, a $90$\,s start gap and both controller choices over five seeds (10 runs). Three-vehicle coordination uses clean Horseshoe Bay, start gaps $8.0$ and $0.0$\,s, seeds 0--2 and matched coordinated and uncoordinated conditions (12 runs); a further six coordinated runs repeat both gaps with the yield margin reduced from two locally estimated gates to one. The three official track layouts are shown in Fig.~\ref{fig:tracks_layout}.

\input{figures/tracks_layout}

The two official controllers are the continuous-servo Continuous Servo (\cmd{rule\_gate\_baseline}) and the staged Center-then-Commit (\cmd{rule\_gate\_center\_then\_commit}). Coordination deliberately forms a heterogeneous convoy: the leader uses Continuous Servo and both followers use Center-then-Commit, while every vehicle maintains its own local tracker. All performance experiments use the reference HoloOcean adapter with a BlueROV2-class vehicle, $\Delta t=0.033$\,s, official onboard-only observations, no current compensation layer, no obstacles and $1.5\times1.5$\,m gate apertures (R4). Time limits are $560$\,s for Horseshoe Bay, current, fleet and coordination runs, $900$\,s for Vertical Serpent and $1300$\,s for Mixed Endurance. Fallback is disabled. Current runs are accepted only when metadata confirms physical \cmd{set\_ocean\_currents} coupling.

Separate setup checks, not controller-performance trials, instantiate both named current profiles and seeded static random obstacles on every official circuit. These checks confirm that the same gate geometry supports medium or strong current and physical obstacle placement; they do not claim that either reference gate follower avoids obstacles.

The reported metrics are course completion, completed gates, official time, penalized time from \eqref{eq:penalized}, gate and world collisions, out-of-bounds events and stuck events. Fleet and coordination validation additionally report inter-vehicle proximity events and team elapsed and penalized times. To keep the statistics unambiguous, \emph{time} statistics are computed over finished runs only, while completed-gate, collision and event means are computed over all seeds; a cell with a single finished run is reported without a spread. Dispersions are sample standard deviations (ddof$=1$). Controller-local progress is compared with independent referee progress for every vehicle. False or missed local advancements, inconsistent final counts and inconsistent finish state fail the progress audit rather than being repaired or hidden; ordinary delay for a matched advancement is reported separately.

\subsection{Single-Vehicle Clean Tracks}
\label{sec:eval_clean}

\input{tables/validation_results}

Table~\ref{tab:validation_results} reports the clean-track results over five seeds. Continuous Servo finishes every Horseshoe Bay and Vertical Serpent run but only $2/5$ Mixed Endurance runs. Center-then-Commit finishes all Horseshoe Bay and Mixed Endurance runs and $4/5$ Vertical Serpent runs. The mean completed-gate counts are respectively $12.0$, $17.0$ and $17.0$ for Continuous Servo and $12.0$, $14.8$ and $22.0$ for Center-then-Commit. No clean run records a gate or world collision, out-of-bounds event or stuck event.

\subsection{Staggered Fleet and Team Scoring}

The bottom block of Table~\ref{tab:validation_results} reports homogeneous two-vehicle fleet validation on Horseshoe Bay over five seeds. With either controller, every team finishes $24/24$ gates without gate or world collisions, inter-vehicle proximity events, out-of-bounds events or stuck events. Mean team elapsed and penalized time are both $303.1\pm3.2$\,s for Continuous Servo and $308.0\pm5.8$\,s for Center-then-Commit. The $90$\,s release gap deliberately produces a low-contact case that validates multi-vehicle spawning, staggered release, independent per-vehicle referee state, ranking and team aggregation for both official controllers; the interacting case is studied with three vehicles below.

\subsection{Behaviour Under Currents}
\label{sec:eval_currents}

Performance degrades as current strength increases. Under the Horseshoe Bay \texttt{medium} profile, Continuous Servo finishes $2/5$ runs, averages $8.4/12$ completed gates and $43.4$ gate and world collisions per run; its finished-only official and penalized times are $337.9\pm19.2$\,s and $692.9\pm75.7$\,s. Center-then-Commit finishes $3/5$, averages $8.8/12$ gates and $25.2$ gate and world collisions, with finished-only times of $217.3\pm1.5$\,s and $223.9\pm1.8$\,s. Neither controller finishes a \texttt{strong} run: Continuous Servo averages $3.0/12$ gates and no gate or world collision, while Center-then-Commit averages $3.2/12$ gates and $0.8$ gate and world collisions. Metadata confirms physical current coupling in every run.

Two features of these numbers deserve comment. First, the collision counts are \emph{non-monotonic} in current strength: the medium profile records many gate and world collisions ($43.4$ and $25.2$) while the strong profile records almost none ($0.0$ and $0.8$). This is not evidence of better control under the stronger current. Under the strong profile the vehicles complete only $3.0$--$3.2$ of $12$ gates and stall early, so they rarely reach the gate-dense regions where contacts accumulate; fewer collisions here reflect less progress, not more skill. Second, the strong profile is best read as a saturation regime. Its constant set-flow component alone is $[0.75,1.05,0]$\,m/s, a magnitude of $1.29$\,m/s, about $2.7\times$ the mean course speed the same vehicle sustains on clean Horseshoe Bay ($93.8$\,m in $194.5$\,s, i.e.\ $\approx\!0.48$\,m/s). The strong profile therefore substantially limits forward course progress and functions as a stress case, whereas the medium profile ($0.65$\,m/s constant component) is the more informative disturbance regime. The quantitative current sweep is limited to Horseshoe Bay (Section~\ref{sec:limitations}).

\subsection{Reference-Controller Comparison}
\label{sec:eval_comparison}

\input{figures/controller_comparison}

The two camera-assisted controllers of Section~\ref{sec:controller} occupy different speed--robustness trade-offs rather than one dominating the other; Fig.~\ref{fig:controller_comparison} summarises their completion across every reported condition. On the short, planar Horseshoe Bay circuit they are effectively tied: both finish all $12$ gates in every seed with no gate or world collision, out-of-bounds event or stuck event, and mean official (and penalized) time is $194.5\pm4.8$\,s for Continuous Servo and $197.7\pm7.1$\,s for Center-then-Commit, so the staged controller is $3.2$\,s ($1.7\%$) slower in this matched case with no measured collision advantage.

The two controllers separate on the harder tracks and under disturbance. On Vertical Serpent, Continuous Servo is more reliable ($5/5$ versus $4/5$ finishes). On the long Mixed Endurance course the ordering reverses and is larger: Center-then-Commit finishes all five seeds and completes every gate ($22.0/22$), whereas Continuous Servo finishes only $2/5$ and averages $17.0/22$ gates. Under the medium current the staged controller is again the more robust of the two, finishing more runs ($3/5$ versus $2/5$), incurring roughly half the collisions ($25.2$ versus $43.4$) and completing its finished runs far faster ($217.3$\,s versus $337.9$\,s). The committed-passage phase, which stops chasing image-centroid changes once locked, thus trades a small amount of clean-track speed for steadier passages when the course is long or the vehicle is pushed off line; the continuous-servo policy retains a slight edge only where constant re-centering is cheap. The comparison is limited to the three official tracks, two current profiles and the evaluated seeds.

\subsection{Three-Vehicle Coordination}
\label{sec:eval_holoocean_coord}

This validation tests local leader-follower coordination under contact dynamics. It uses clean Horseshoe Bay in official mode with three vehicles, no currents or obstacles, two start gaps ($8.0$ and $0.0$\,s) and three seeds. The leader runs Continuous Servo and both followers run Center-then-Commit; every controller advances its own LocalCourseTracker and heartbeats carry only local estimates. No controller or course parameter is retuned between conditions. We report the recommended one-gate margin LF(1), the conservative two-gate margin LF(2) and the matched uncoordinated baseline (Table~\ref{tab:holoocean_coordination}).

Coordination removes physical interaction that the uncoordinated convoy suffers, and the recommended LF(1) margin does so at the lowest time cost. At an $8$\,s gap, the uncoordinated convoy finishes only $2/3$ runs, averages $34.3/36$ team gates, $127.3$ gate and world collisions and $2.0$ inter-vehicle proximity events per run; LF(1) and LF(2) both finish $3/3$ with no gate, world or proximity events, and LF(1) is faster ($266.0\pm8.3$\,s versus $285.7\pm3.8$\,s team time). At simultaneous release, the uncoordinated convoy still finishes $3/3$ but averages $9.3$ gate and world collisions and $2.3$ inter-vehicle proximity events; LF(1) again eliminates both event types at $262.3\pm7.7$\,s, while the more conservative LF(2) is slower ($288.3\pm1.3$\,s raw, $303.3\pm1.3$\,s penalized) and, because its longer hold can trip the independent stuck timer, records one stuck event per run. Across both gaps LF(1) finishes every run with zero gate, world, proximity, out-of-bounds and stuck events and is faster than LF(2); LF(2) keeps a larger nominal gate separation but did not improve safety in these conditions, which is why LF(1) is the recommended default and LF(2) is retained only as a conservative comparison. This evidence is limited to one clean track, three vehicles, two gaps and three seeds (Section~\ref{sec:limitations}).

\input{tables/holoocean_coordination}

\subsection{Reproducibility}
\label{sec:reproducibility}

Every run emits a structured event log and summary. The repository provides the exact commands used here~\cite{holodronecompetition2026}, together with an artifact manifest that records, per run, the experiment-generation commit, the source-tree fingerprint, the HoloOcean environment, the track, controller, seed, command, fallback-disabled status and physical current-coupling status. A single configuration-driven entry point and per-track JSON files make each official run reproducible from one command, and the single-vehicle, fleet and coordination sweeps are scripted, so the HoloOcean result tables can be regenerated from the stored artifacts by \path{article/regenerate_tables.py} without rerunning the simulator.
