\section{Conclusion}
\label{sec:conclusion}

This paper presented \mra{}, a configurable benchmark layer that isolates vehicle-specific and simulator-specific behavior behind a replaceable adapter; the reported implementation uses HoloOcean and a BlueROV2-class agent. It defines single-vehicle evaluation, then extends it to staggered cooperative fleets with team scoring. Controllers see only the official observation, while an independent referee uses privileged state for crossings, penalties, timeouts, ranking and scores. We formalized the benchmark instance, gate geometry, single-rover and team scoring and staggered fleet semantics, then grounded them in the implementation.

Across the five-seed clean validation, continuous servo finishes $5/5$ Horseshoe Bay, $5/5$ Vertical Serpent and $2/5$ Mixed Endurance runs; center-then-commit finishes $5/5$, $4/5$ and $5/5$ respectively. Both homogeneous two-rover fleet variants finish all five seeds without gate/world collisions, inter-vehicle proximity events, out-of-bounds events or stuck events. Under the \texttt{medium} current profile, continuous servo finishes $2/5$ and center-then-commit $3/5$; neither controller finishes a \texttt{strong} run. These failures remain benchmark evidence because gate geometry, referee margins and current profiles are unchanged.

In the matched clean Horseshoe Bay comparison, both camera-assisted controllers finish every seed without collision; continuous servo averages $194.456$\,s and center-then-commit $197.683$\,s, so this case supports neither a speed nor a collision advantage for staged passage. In the three-rover diagnostic, the primary leader--follower policy finishes all six runs and eliminates gate/world collisions and inter-vehicle proximity events, although simultaneous release produces one stuck event per seed. Without coordination, one $8$\,s-gap team fails and contact costs are substantial. The one-gate-margin ablation finishes all six runs cleanly and more quickly than the primary margin in this limited sample. Execution and artifact-contract checks pass with zero failures, but the overall onboard-only audit fails on 15 runs with controller-local/referee progression disagreement; these outcomes are preserved rather than concealed.
